---
title: "Lab 1 - DYI fitting MLM"
author: "Alencar Xavier, Gota Morota"
date: "October 26, 2018"
widescreen: true
output: ioslides_presentation
---

# Example 1 - Sorghum


## Example 1 - load data

Example dataset from Kevin's *agridat* package

```{r}
data(adugna.sorghum, package = 'agridat')
dt = adugna.sorghum
head(dt)
```

## Example 1 - Getting design matrix

- Linear model: $Pheotype = Env + Gen$
- In algebra notation: $y = Xb + Zu + e$

```{r}
y = dt$yield
X = model.matrix(y~env,dt)
Z = model.matrix(y~gen-1,dt) # "-1" means no intercept
```

Assuming:

- $u\sim N(0,I\sigma^2_g)$
- $e\sim N(0,I\sigma^2_e)$

## Example 1 - Visualize X and Z matrices

```{r,echo=T,fig.width=8,fig.height=4,fig.align="center"}
SEE=function(A,...)image(t(1-A[nrow(A):1,]),axes=F,col=gray.colors(2),...)
par(mfrow=c(1,2),mar=c(1,1,3,1))
SEE(X, main=paste("X matrix (",paste(dim(X),collapse=' x '),")" ))
SEE(Z, main=paste("Z matrix (",paste(dim(Z),collapse=' x '),")" ))
```

## Example 1 - Fit the model

```{r}
# Using the NAM package (same for rrBLUP, EMMREML, BGLR)
require(NAM, quietly = TRUE)
fit1 = reml(y=y,X=X,Z=Z)

# Alternatively, you can also use formulas with NAM
fit1b = reml(y=dt$yield,X=~dt$env,Z=~dt$gen )

# Using the lme4 package
require(lme4, quietly = TRUE)
fit2 = lmer(yield ~ env + (1|gen), data=dt)
```

## Example 1 - Variance components

```{r}
fit1$VC[c(1:2)] # same with fit1b$VC
data.frame((summary(fit2))$varcor)$vcov
```

- VC can be used to measure *broad-sense* heritability

$H=\frac{\sigma^2_g}{\sigma^2_g+\sigma^2_e/n}=\frac{189680.4}{189680.4+442075.6/10.32}=0.82$

## Example 1 - The coefficients

```{r,fig.width=8,fig.height=4,fig.align="center"}
blue1 = fit1$Fixed[,1];  blup1 = fit1$EBV
blue2 = fit2@beta;       blup2 = rowMeans(ranef(fit2)$gen)
par(mfrow=c(1,2));
plot(blue1,blue2,main="BLUE"); plot(blup1,blup2,main="BLUP")
```

## Example 1 - DYI BLUPs

```{r}
iK = diag(ncol(Z))
Lambda = 442075.6/189680.4
W = cbind(X,Z)
Sigma = as.matrix(Matrix::bdiag(diag(0,ncol(X)),iK*Lambda))
LHS = crossprod(W) + Sigma
RHS = crossprod(W,y)
g = solve(LHS,RHS)
my_blue = g[ c(1:ncol(X))]
my_blup = g[-c(1:ncol(X))]
```

## Example 1 - DYI BLUPs

```{r,fig.width=8,fig.height=4,fig.align="center"}
par(mfrow=c(1,2))
plot(my_blue,blue1,main="BLUE")
plot(my_blup,blup1,main="BLUP")
```

## Example 1 - DYI Variance components

$\sigma^{2}_{e} = \frac{e'y}{n-p}$ and $\sigma^{2}_{u} = \frac{u'K^{-1}u+tr(K^{-1}C^{22}\sigma^{2}_{e})}{q}$

```{r}
e = y - X %*% my_blue - Z %*% my_blup
Ve = c(y%*%e)/(length(y)-ncol(X))
Ve
trKC22 = sum(diag(iK%*%(solve(LHS)[-c(1:ncol(X)),-c(1:ncol(X))])))
Vg = Vg = c(t(my_blup)%*%iK%*%my_blup+trKC22*Ve)/ncol(Z)
Vg
```


## Starting from bad variance components

```{r}
Ve = Vg = 1
for(i in 1:25){
  Lambda = Ve/Vg; 
  Sigma = as.matrix(Matrix::bdiag(diag(0,ncol(X)),iK*Lambda))
  LHS = crossprod(W) + Sigma; RHS = crossprod(W,y); g = solve(LHS,RHS)
  my_blue = g[ c(1:ncol(X))]; my_blup = g[-c(1:ncol(X))]
  e = y - X%*%my_blue - Z%*%my_blup; Ve = c(y%*%e)/(length(y)-ncol(X))
  trKC22 = sum(diag(iK%*%(solve(LHS)[(ncol(X)+1):(ncol(W)),(ncol(X)+1):(ncol(W))])))
  Vg = c(t(my_blup)%*%iK%*%my_blup+trKC22*Ve)/ncol(Z)
  if(!i%%5){cat('It',i,'VC:  Vg =',Vg,'and Ve =',Ve,'\n')}}
```

# Example 2 - Barley

## Example 2 - load data

Another example dataset from Kevin's *agridat* package

```{r}
data(steptoe.morex.pheno,package='agridat')
dt = steptoe.morex.pheno
head(dt)
```

## Example 2 - Getting design matrix

- Linear model: $Phe=Env+Gen$

- In algebra notation: $y=Xb+Zu+e$  

```{r}
X = model.matrix(~env,dt)
Z = model.matrix(~gen-1,dt) # "-1" means no intercept
y = dt$yield
```

## Example 2 - Fit the model

```{r}
# Fit
fit0 = reml(y=y,X=X,Z=Z)

# BLUE and BLUP
blue0 = fit0$Fixed[,1]
blup0 = fit0$EBV

# Get VC
fit0$VC[c(1:2)]
```

## Example 2 - DYI BLUPs

```{r}
iK = diag(ncol(Z))
Lambda = 0.637997/0.132009
W = cbind(X,Z)
Sigma = as.matrix(Matrix::bdiag(diag(0,ncol(X)),iK*Lambda))
LHS = crossprod(W) + Sigma
RHS = crossprod(W,y)
g = solve(LHS,RHS)
my_blue = g[ c(1:ncol(X))]
my_blup = g[-c(1:ncol(X))]
```

## Example 2 - DYI BLUPs

```{r,fig.width=8,fig.height=4,fig.align="center"}
par(mfrow=c(1,2))
plot(my_blue,blue0,main="BLUE")
plot(my_blup,blup0,main="BLUP")
```

## Example 2 - Check variance components

$\sigma^{2}_{e} = \frac{e'y}{n-p}$ and $\sigma^{2}_{u} = \frac{u'K^{-1}u+tr(K^{-1}C^{22}\sigma^{2}_{e})}{q}$

```{r}
e = y - X %*% my_blue - Z %*% my_blup
Ve = c(y%*%e)/(length(y)-ncol(X))
Ve
trKC22 = sum(diag(iK%*%(solve(LHS)[-c(1:ncol(X)),-c(1:ncol(X))])))
Vg = c(t(my_blup)%*%iK%*%my_blup+trKC22*Ve)/ncol(Z)
Vg
```

## Starting from bad variance components

```{r}
Ve = Vg = 1
for(i in 1:25){
  Lambda = Ve/Vg; 
  Sigma = as.matrix(Matrix::bdiag(diag(0,ncol(X)),iK*Lambda))
  LHS = crossprod(W) + Sigma; RHS = crossprod(W,y); g = solve(LHS,RHS)
  my_blue = g[ c(1:ncol(X))]; my_blup = g[-c(1:ncol(X))]
  e = y - X%*%my_blue - Z%*%my_blup; Ve = c(y%*%e)/(length(y)-ncol(X))
  trKC22 = sum(diag(iK%*%(solve(LHS)[(ncol(X)+1):(ncol(W)),(ncol(X)+1):(ncol(W))])))
  Vg = c(t(my_blup)%*%iK%*%my_blup+trKC22*Ve)/ncol(Z)
  if(!i%%5){cat('It',i,'VC:  Vg =',Vg,'and Ve =',Ve,'\n')}}
```

# Example 3 - Barley GEBV

## Using genomic information!

```{r}
data(steptoe.morex.geno,package='agridat')
gen = do.call("cbind",lapply(steptoe.morex.geno$geno,function(x) x$data))
gen = rbind(0,2,gen)
rownames(gen) = c('Morex','Steptoe',as.character(steptoe.morex.geno$pheno$gen))
rownames(gen)[10] = "SM8"
gen = gen[gsub('gen','',colnames(Z)),]
K = G2A_Kernels(gen)$A
```

## Example 3 - Fit the model

```{r}
# Fit model
fit0 = reml(y=y,X=X,Z=Z,K=K)

# BLUE and BLUP
blue0 = fit0$Fixed[,1]
gebv0 = fit0$EBV

# Get VC
fit0$VC[c(1:2)]
```

## Example 3 - DYI BLUPs

```{r}
iK = chol2inv(K)
Lambda = 0.6575786/0.2334026
W = cbind(X,Z)
Sigma = as.matrix(Matrix::bdiag(diag(0,ncol(X)),iK*Lambda))
LHS = crossprod(W) + Sigma
RHS = crossprod(W,y)
g = solve(LHS,RHS)
my_blue = g[ c(1:ncol(X))]
my_gebv = g[-c(1:ncol(X))]
```

## Example 3 - DYI BLUPs

```{r,fig.width=8,fig.height=4,fig.align="center"}
par(mfrow=c(1,2))
plot(my_blue,blue0,main="BLUE")
plot(my_gebv,gebv0,main="GEBV")
```

## Example 3 - Check variance components

$\sigma^{2}_{e} = \frac{e'y}{n-p}$ and $\sigma^{2}_{u} = \frac{u'K^{-1}u+tr(K^{-1}C^{22}\sigma^{2}_{e})}{q}$

```{r}
e = y - X %*% my_blue - Z %*% my_blup
Ve = c(y%*%e)/(length(y)-ncol(X))
Ve
trKC22 = sum(diag(iK%*%(solve(LHS)[(ncol(X)+1):(ncol(W)),(ncol(X)+1):(ncol(W))])))
Vg = c(t(my_blup)%*%iK%*%my_blup+trKC22*Ve)/ncol(Z)
Vg
```

# Example 4 - Soybeans

## snp-BLUP

```{r}
data(tpod,package='NAM')
X = matrix(1,length(y),1)
Z = gen
dim(Z)
```

## Example 3 - Fit the model

```{r,fig.width=8,fig.height=3.5,fig.align="center"}
# Fit using the lme4 package
fit0 = reml(y=y,X=X,Z=Z) # same as reml(y=y,Z=gen) 
marker_values = fit0$EBV
gebv0 = c(gen %*% marker_values)
# Marker effects
plot(marker_values,pch=16, xlab='SNP')
```

## Example 3 - DYI BLUPs

```{r}
iK = diag(ncol(Z))
Lambda = fit0$VC[2] / fit0$VC[1]
W = cbind(X,Z)
Sigma = diag( c(0,rep(Lambda,ncol(Z))) )
LHS = crossprod(W) + Sigma
RHS = crossprod(W,y)
g = solve(LHS,RHS)
my_mu = g[ c(1:ncol(X))]
my_marker_values = g[-c(1:ncol(X))]
my_gebv = c(gen %*% my_marker_values) # GEBVs from RR
```

## Example 3 - DYI BLUPs

```{r,fig.width=8,fig.height=4,fig.align="center"}
par(mfrow=c(1,2))
plot(my_marker_values, marker_values, main="Markers")
plot(my_gebv, gebv0, main="GEBV")
```

## Example 3 - Heritability from RR

```{r,fig.width=4,fig.height=4,fig.align="center"}
fit0$VC

Scale=sum(apply(gen,2,var)); Va=fit0$VC[1]*Scale;  Ve=fit0$VC[2]
round((Va/(Va+Ve)),2)

K = tcrossprod(apply(gen,2,function(x) x-mean(X)))
K = K/mean(diag(K)); round(reml(y,K=K)$VC,2)
```

## Estimate VC from bad starters

```{r}
W = cbind(X,Z); iK = diag(ncol(Z))
Ve = Vg = 1 # Bad starting values
for(i in 1:100){ # Check the VC convergence after few iterations
  Lambda = Ve/Vg; 
  Sigma = as.matrix(Matrix::bdiag(diag(0,ncol(X)),iK*Lambda))
  LHS = crossprod(W) + Sigma; RHS = crossprod(W,y); g = solve(LHS,RHS)
  my_blue = g[ c(1:ncol(X))]; my_blup = g[-c(1:ncol(X))]
  e = y - X%*%my_blue - Z%*%my_blup; Ve = c(y%*%e)/(length(y)-ncol(X))
  trKC22 = sum(diag(iK%*%(solve(LHS)[(ncol(X)+1):(ncol(W)),(ncol(X)+1):(ncol(W))])))
  Vg = c(t(my_blup)%*%iK%*%my_blup+trKC22*Ve)/ncol(Z)
  if(!i%%25){cat('It',i,'VC:  Vg =',Vg,'and Ve =',Ve,'\n')}}
```